### 2018-03-20 submission/manuscript.Rmd
add two more paragraphs to discussion

>>>>

Similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community. The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. However, the chimarea prevalence for some HiFi DNA polymerases (e.g. Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study. Within the larger context of the different 16S rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used.

Heavy standardization has been commonly suggested as a reasonable answer to finding the most reproducible approach. However, bias can be easily reproducible and can be found in every step of the 16S rRNA gene sequencing workflow. This study shows that the observed microbial community consistently varies based on HiFi DNA polymerase. Standardizing multiple workflows to one specific HiFi DNA polymerase could be detrimental since the PCR step not only misses entire species based on primer chosen () but also can artifically increase the number of OTUs observed. Arguably, standardizing workflow needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

### 2018-03-20 submission/manuscript.tex
add two more paragraphs to discussion

>>>>

Similar to using different perservation methods or different DNA
extraction kits, the type of HiFi DNA polymerase can add bias to the
observed bacterial community. The sequence error introduced by the HiFi
DNA polymerase is small and likely to be smaller than the biological
variation within a specific study, which would be consistent with
previous findings for preservation and DNA extraction methods (Salter et
al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
large and might be greater than the oberseved biological variation
within a specific study. Within the larger context of the different 16S
rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as
important a consideration as either preservation or DNA extraction
method used.

Heavy standardization has been commonly suggested as a reasonable answer
to finding the most reproducible approach. However, bias can be easily
reproducible and can be found in every step of the 16S rRNA gene
sequencing workflow. This study shows that the observed microbial
community consistently varies based on HiFi DNA polymerase.
Standardizing multiple workflows to one specific HiFi DNA polymerase
could be detrimental since the PCR step not only misses entire species
based on primer chosen () but also can artifically increase the number
of OTUs observed. Arguably, standardizing workflow needs to be
approached on a study by study basis and not every project needs to use
the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.
